A Parallel Geometric Multigrid Algorithm

### 2014-09-24 branches/testcharge/htdocs/binary.tex
sync merge trunk to testcharge
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2478 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  sigma.tex
%
%  A rather long section on the calibration of surface tension
%  for the binary fluid models.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Calibration of the Surface Tension}

The binary fluid model introduces an interfacial tension $\sigma$ which
can be predicted from the choice of free energy \cite{swift,kendon}.
In particular, Kendon et al. 2000 described the process of calibrating
the actual interfacial tension of the model fluid. This was based on
a measurement of the Laplace pressure difference between the fluid
inside and outside a spherical droplet via
\begin{equation}
\Delta p = 2 \sigma / R
\end{equation}
where $R$ is the radius of the spherical drop at equilibrium. This
measurement appears quite difficult in practice owing to tendency of
any initial drop to evaporate \cite{yue2007}.

\subsection{New measurement}

Here, we prefer a measure based on the expected profile of the
order parameter $\phi$ and the free energy density at a flat interface.
In order to sample all orientations of the interface with respect to
the lattice we again take a large droplet of fluid A initialised at
rest in fluid B. We assume that the interface is locally flat and so
the procedure holds.

If a droplet with radius $r_0$ is constructed with an initial interfacial
profile of $\phi(r) = \tanh{(r - r_0) / \xi_0}$
with $r$ in the radial direction,  then the free energy density will be
(with $-A = B$)
\begin{equation}
 e(r) = -{\scriptstyle \frac{1}{4}} |A| [1 - \mathrm{sech}^4 (r-r_0)/\xi_0]
+ {\scriptstyle \frac{1}{2}}
 (\kappa/\xi_0^2)\, \mathrm{sech}^4{(r-r_0)/\xi_0}.
\end{equation}
Obtaining a fit to $\xi_0$ from the measured interfacial profile of
$\phi (r)$ and to $ e_0 = \kappa / 2 \xi_0^2$ from the peak in the measured
profile of free energy
density allows an estimate of the interfacial tension to be made:
\begin{equation}
\sigma = 2(8/9)^{1/2} \xi_0 |e_0|.
\end{equation}
This measurement is taken after the initial profile has been allowed
to relax for at least 10 times the larger of the diffusion time
$\xi_0^2 / M$ or the momentum diffusion time $\xi_0^2 / \eta$.

We have measured the surface tension using the following approaches.

\textit{Method 1}.

LB for two distributions with relaxation of order parameter
following Kendon etal. The force on the fluid is applied
via the equilibrium stress following Swift et al.

\textit{Method 2}.

As method 1 but using the realxation of order parameter
following Stratford etal.

\textit{Method 3}.

As Method 1 but with force via divergence of the chemical stress.

\textit{Method 4}.

As Method 3 but with force via divergence of the chemical stress.

\textit{Method 5}.

Using finite difference (1st order upwind) for the order parameter.

\textit{Method 6}.

Using finite difference (3rd order upwind).

\subsection{Results}

We follow the parameter sets in Kendon et al Table~2. These
are reproduced here in Table~\ref{tab:r1}. Measured values for
Run028 parameters are show in Table~\ref{tab:newr028}.

\begin{table}
\begin{center}
\begin{tabular}{llllllll}
\hline
Run & $-A, B$ & $\kappa$ & $\eta$ & $M$ & $\sigma$  & $L_0$ & $T_0$ \\
\hline
Rks001 & 0.00625 & 0.004 & 1.41 & 0.05 & 0.0047 & 422 & 1.26e10$^5$\\
Rks002 & 0.00625 & 0.004 & 0.65 & 0.25 & 0.0047 & 89.7 & 1.24e10$^4$\\
Run028 & 0.083 & 0.053 & 1.41 & 0.05  & 0.0625 & 31.8 & 718 \\
Run022 & 0.0625 & 0.04 & 0.5 & 0.25   & 0.0471 & 5.31 & 56.3 \\
Run029 & 0.0625 & 0.04 & 0.2 & 0.15   & 0.0471 & 0.849 & 3.61 \\
Run020 & 0.00625 & 0.004 & 0.025 & 2.00 & 0.00471 & 0.133 & 0.704 \\
Run030 & 0.00625 & 0.004 & 0.0065 & 1.25 & 0.00471 & 0.00829 & 0.0124 \\
Run019 & 0.003125 & 0.002 & 0.0014 & 4.00 & 0.00236 & 0.000831 & 0.000493 \\
Run032 & 0.00125  & 0.0008 & 0.0005 & 5.0 & 0.000943 & 0.000265 & 0.000141\\
\hline  
\end{tabular}
\end{center}
\caption{Table showing common parameter values following Kendon et al.
with the theoretical interfacial tension, $L_0$ and $T_0$. Note that
the order parameter mobility is used here (cf $\tilde{M} = 2M$) in
Kendon et al. (KS values are extra sets used to low 1/$\dot{\gamma}T_0$.)}
\label{tab:r1}
\end{table}

\begin{table}
\begin{tabular}{lllllll}
\hline
Method & $\xi_0$ & $e_{min}$ & $e_{max}$ & $\sigma$  & $\phi_{min}$
& $\phi_{max}$\\
\hline
\multicolumn{7}{c}{Run028}\\
\hline
Theory   & 1.13 & -0.0207 &  0.0207 & 0.0625 & -1.0 & 1.0 \\
Method 1 & 1.00 & -0.0207 & 0.0128 & 0.063(3) & -1.0091 & 1.0028 \\
Method 2 & 1.00 & -0.0207 & 0.0126 & 0.063(1) & -1.0113 & 0.9993 \\
Method 3 & && & & & \\
Method 4 & && & & & \\
Method 5 & && & & & \\
Method 6 & --& --& --& --& --& --\\
\hline
\multicolumn{7}{c}{Run029}\\
\hline
Theory & 1.13 & -0.0156 & 0.0156 & 0.0471  & -1.0 & 1.0\\
Method 2 & 0.99 & -0.0156 & 0.010(?) & 0.048(4) & -1.0110 & 1.0003\\ 
Method 6 & 1.02 & -0.0156 & 0.0099 & 0.049(0) & -1.0048 & 1.0002\\
\hline
\multicolumn{7}{c}{Run030}\\
\hline
Theory & 1.13 & -0.00156 & 0.00156 & 0.00471 & -1.0 & 1.0\\
Method 2 & 1.04 & -0.00156 & 0.0008(?) & 0.0047(1) & -1.0167 & 1.0063\\
Method 6 & 1.08 & -0.00156 & 0.0007(?) & 0.0048(0) & -1.0028 & 1.0031\\
\hline
\multicolumn{7}{c}{Run019}\\
\hline
Theory & 1.13 & -0.000781 & 0.000781 & 0.00236 & -1.0 & 1.0\\
Method 2 & 1.29(?) & -0.000781 & 0.00015(?) & 0.0022(8) & -1.0312 & 1.0242\\
Method 6 & 1.23(?) & 0.0010(?) &&  0.0024(?) & -1.0085 & 1.0019\\
\hline 
\multicolumn{7}{c}{Run032}\\
\hline
Theory   & 1.13 &  -0.000312 & 0.000312 & 0.000943  & -1.0 &1.0\\
Method 2 & 1.07  & -0.000312& 0.000157 & 0.00094(7) & -1.0511 & 1.0311\\
Method 6 & 1.6(5)& -0.000312 & not st. & 0.0010(7) &-1.0056 & 1.0010\\
\hline
\end{tabular}
\caption{Results for the different methods for Run208 parameters.}
\end{table}

\section{Upwind Advection Schemes}

@@

\subsection{Uniformly third order scheme}

@@

\subsection{Gory details}

@@

\section{Lattice kinetic equation viewed as finite difference}

@@

\subsection{Advective terms}

@@

\subsection{Dissipative terms}

@@

\subsection{Chemical potential term}

@@

>>>>

\section{Binary Fluids}

\subsection{The Cahn-Hilliard Equation}

\subsection{The Choice of Free Energy}

\subsection{A Second Distribution Function}

\subsection{The Collision}

\subsection{Bounce-Back on Links}

\subsection{Upwind Advection Schemes}

@@

\subsubsection{Uniformly third order scheme}

@@

\subsubsection{Gory details}

@@

\subsection{Lattice kinetic equation viewed as finite difference}

@@

\subsubsection{Advective terms}

@@

\subsubsection{Dissipative terms}

@@

\subsubsection{Chemical potential term}

@@

\subsection{User input}

\inputkey{free\_energy symmetric}

The following applies for the binary fluid problem with
compositional order parameter $\phi$ and free energy
(excluding the term in the density $\rho$):
\begin{equation}
 F[\phi] = 
\int dr \left(
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right).
\end{equation}

This is described in some detail
by Kendon et al \cite{viv}. The first two terms represent the bulk
contribution, whereas the term in $\kappa$ penalises curvature in
the interface.

\inputkey{A} The parameter $A$. Note $A < 0$.

\inputkey{B} The parameter $B$. Note $B = -A$ for common usage,
although this is not enforced.

\inputkey{K} The parameter $\kappa$, which is positive.

The order parameter evolution is determined by a Cahn-Hilliard equation
with mobility set by

\inputkey{mobility} Sets mobility $M$ (uniform in space).

The following parameters control the initialisation of the order
parameter.

\inputkey{phi\_initialisation}

Determines how the compositional order parameter is initialised
at the start of the run. If set to \texttt{spinodal} the value
is set to $\phi_0$ as set above plus or minus a random noise, the
magnitude of which is set by the value of the \texttt{noise} key.
If set to \texttt{block} a one-dimensional profile is set up in
the $z$-direction representing two blocks of fluid with
$\phi = \pm 1$. The two interfaces are set at $z = L_z/4$ and
$z = 3L_z/4$ with the equilibrium $\tanh(z/\xi_0)$ profile having 
appropriate width. The $\phi = -1$ section is in the middle.

\inputkey{phi0}

The mean compositional order parameter roughly $-0.5 < \phi_0 < 0.5$.
The default value is zero, i.e., a symmetric 50:50 mixture by volume.

\inputkey{noise}

The magnitude of the initial fluctuations in $\phi$ used to
initiate spinodal decomposition.

\subsubsection{Binary fluid using two distributions}

\inputkey{symmetric\_lb}

This is the special case where the composition is represented
by a second LB distribution, and an appropriate lattice kinetic
equation approximating the Cahn-Hilliard equation is solved.
In this case the above parameters have the same meaning.

Note that the relaxation time for the second distribution is
related to the mobility by $tau_\phi = M\rho_0/\Delta t + 1/2$.

\subsubsection{Brazovskii}

\inputkey{free\_energy brazovskii}

This is similar to the symmetric free energy, but with one extra term
in a higher derivative of $\phi$.
\begin{equation}
 F[\phi] = 
\int dr \left(
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
+ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
\end{equation}

The parameters now include $C$. For $A<0$, phase separation occurs
with a result depending on $\kappa$: ones get two symmetric phases
for $\kappa >0$ (cf. symmetric) or a lamellar phase for
$\kappa < 0$.

\inputkey{A} Bulk parameter $A < 0$.

\inputkey{B} Bulk parameter $B = -A$.

\inputkey{K} Negative for lamellar phase.

\inputkey{C} Positive.

Other parameters, including the mobility, are set as for the
symmetric free energy (see above).
